Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-04 inference.tex

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds one to $|\nubar|$, if $s$ is a bind expression nested
inside a top-level bind expression (like $\F_U(\stl,\stg)$ from
the above example), then the prenex signature of $\phi_s$ is
$\forall^2\exists$.  Furthermore, if the subexpressions of $s$ are
neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi =
\bigwedge_i\phi_i \wedge \phi_s$ remains $\forall^2\exists$. Similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside the
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

The fragment of FOL that admits formulas with prenex signatures of
form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
GKS}) fragment~\cite{gks}, which is known to be decidable. The
language of encoding, however, is a combination of {\sf GKS} with (a).
$\Prop^0$, the theory from which quantifier-free propositions
($\varphi$) that encode object language expressions are drawn, and
(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus
the encoding of the subset of $\SL$ described above is decidable if the
combination of ${\sf GKS} + \Prop^0 + \Prop^1$ is decidable. We write
$\SL[\Prop^0,\Prop^1]$ to highlight the parameterization of $\SL$ on
$\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
points to the existence of subsets in $\SL[\Prop^0,\Prop^1]$ that are
decidable:
\begin{theorem}
  There exists an $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
  such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
An example of such an $\SL'$ is the subset described above: $\SL$ with
bind expressions confined to one level of nesting. We denote this
subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
\begin{theorem}
  $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
$\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
programs without nested \C{foreach} loops. Observe that the
\C{new\_order} transaction (Fig.~\ref{fig:new_order_code}) belongs to
this subset.  Indeed, $\SL^1$, while being a restricted version of
$\SL$, is nonetheless expressive enough to cover all the benchmarks we
considered in Sec.~\ref{sec:case-studies}.

A useful instantiation of $\SL^1$ is $\SL[{\sf BV},{\sf GKS}+{\sf
BV}]$, where ${\sf BV}$ is the theory of bit-vector arithmetic, which
is often used to encode the finite-bit integer arithmetic of real
programs. Finite-bit integer arithmetic has a finite axiomatization in
{\sf GKS}. For instance, 32-bit integers can be encoded as $2^{32}$
distinct constants of an uninterpreted sort $T$, while integer
operations like addition and multiplication can be encoded as
uninterpreted functions whose properties are enumerated for the entire
domain of $T$. Thus {\sf BV} is subsumed by {\sf GKS}. Since the
latter is decidable, the combination is decidable:
\begin{theorem}
  $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ is decidable.
\end{theorem}
The $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ instantiation requires $I$ to
be drawn from {\sf GKS}+{\sf BV}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in \S\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf GKS}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable.

>>>>

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds a quantifier to $|\nubar|$, if $s$ is a bind
expression nested inside a top-level bind expression (like
$\F_U(\stl,\stg)$ from the above example), then the prenex signature
of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions
of $s$ are neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi = \bigwedge_i\phi_i \wedge \phi_s$
remains $\forall^2\exists$. A similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside a
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

The fragment of FOL that admits formulas with prenex signatures of the
form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
GKS}) fragment~\cite{gks}, which is known to be decidable. The
language of encoding, however, is a combination of {\sf GKS} with (a).
$\Prop^0$, the theory from which quantifier-free propositions
($\varphi$) that encode object language expressions are drawn, and
(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus,
the encoding of the subset of $\SL$ described above is decidable if the
combination of ${\sf GKS} + \Prop^0 + \Prop^1$ is decidable. We write
$\SL[\Prop^0,\Prop^1]$ to highlight the parameterization of $\SL$ on
$\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
points to the existence of non-trivial subsets in $\SL[\Prop^0,\Prop^1]$ that are
decidable:
\begin{theorem}
  There exists an $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
  such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
One interesting example of such an $\SL'$ is the subset described above: $\SL$ with
bind expressions confined to one level of nesting. We denote this
subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
\begin{corollary}
  $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{corollary}
$\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
programs without nested \C{foreach} loops. Observe that the
\C{new\_order} transaction (Fig.~\ref{fig:new_order_code}) belongs to
this subset.  Indeed, $\SL^1$, while being a restricted version of
$\SL$, is nonetheless expressive enough to cover all the benchmarks we
considered in Sec.~\ref{sec:case-studies}.

A useful instantiation of $\SL^1$ is $\SL[{\sf BV},{\sf GKS}+{\sf
BV}]$, where ${\sf BV}$ is the theory of bit-vector arithmetic, which
is often used to encode the finite-bit integer arithmetic of real
programs. Finite-bit integer arithmetic has a finite axiomatization in
{\sf GKS}. For instance, 32-bit integers can be encoded as $2^{32}$
distinct constants of an uninterpreted sort $T$, while integer
operations like addition and multiplication can be encoded as
uninterpreted functions whose properties are enumerated for the entire
domain of $T$. Thus {\sf BV} is subsumed by {\sf GKS}. Since the
latter is decidable, the combination is decidable:
\begin{theorem}
  $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ is decidable.
\end{theorem}
This instantiation requires $I$ to be drawn from {\sf GKS}+{\sf BV},
which is expressive enough to describe common database integrity
constraints, such as referential integrity, non-negativeness of all
integer values in a column etc.  The isolation specifications
presented in \S\ref{sec:isolation} are already simple first-order
formulas that can be encoded in {\sf GKS}.  Furthermore, it is also
reasonable to expect the guarantee ($G$) of a transaction to be
expressible in the same logic as its inferred $\F$, since $\F$
(without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable.
